---
title: "CSC487-HW1"
author: "Corey Malone"
date: "2025-01-29"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

############################################################
# 1. Work with "Su_raw_matrix.txt" 
############################################################

```{r}
library(ggplot2)

# (a) Read data using read.delim
#    'header=TRUE' means we assume the first row is column names.
su <- read.delim("Su_raw_matrix.txt", header = TRUE)

# Let's quickly check the structure (optional)
str(su)
```

```{r}
# (b) Compute mean and standard deviation of Liver_2.CEL column
mean_Liver2 <- mean(su$Liver_2.CEL, na.rm = TRUE)  # na.rm=TRUE to ignore any NA values
sd_Liver2   <- sd(su$Liver_2.CEL, na.rm = TRUE)

# Print results
cat("Mean of Liver_2.CEL:", mean_Liver2, "\n")
cat("SD of Liver_2.CEL:", sd_Liver2, "\n")

# (c) Get average (colMeans) and total (colSums) values for each column
column_means <- colMeans(su, na.rm = TRUE)
column_sums  <- colSums(su, na.rm = TRUE)

cat("\nColumn means:\n")
print(column_means)

cat("\nColumn sums:\n")
print(column_sums)
```

############################################################
# 2. Generate random numbers from normal distributions and plot histograms 
############################################################

We'll generate 10,000 random values from N(0, 0.2^2) and N(0, 0.5^2)
Then plot histograms and comment on their differences.

```{r}
# (a) mean=0, sigma=0.2
set.seed(123)  # Optional, for reproducible results
x1 <- rnorm(10000, mean = 0, sd = 0.2)

hist(x1,
     main = "Histogram of N(0,0.2^2)",
     xlab = "Values",
     xlim = c(-5,5),   # so we can compare easily
     breaks = 50,
     col = "lightblue")
```

```{r}
# (b) mean=0, sigma=0.5
set.seed(123)
x2 <- rnorm(10000, mean = 0, sd = 0.5)

hist(x2,
     main = "Histogram of N(0,0.5^2)",
     xlab = "Values",
     xlim = c(-5,5),
     breaks = 50,
     col = "lightpink")
```
# Comment on differences:
 The distribution with sigma=0.2 is narrower (less spread),
 while sigma=0.5 is wider (more spread). Both have mean 0.


############################################################
# 3. Use ggplot2 with "dat" and then "diabetes" 
############################################################

First, create a sample dataframe 'dat'
```{r}
dat <- data.frame(
  cond = factor(rep(c("A", "B"), each = 200)),
  rating = c(rnorm(200), rnorm(200, mean = 0.8))
)
```

# (b) Overlaid histograms
```{r}
ggplot(dat, aes(x = rating, fill = cond)) +
  geom_histogram(binwidth = 0.5, alpha = 0.5, position = "identity") +
  ggtitle("Overlaid Histograms (dat)")
```

# (c) Interleaved histograms
```{r}
ggplot(dat, aes(x = rating, fill = cond)) +
  geom_histogram(binwidth = 0.5, position = "dodge") +
  ggtitle("Interleaved Histograms (dat)")
```

# (d) Density plots
```{r}
ggplot(dat, aes(x = rating, colour = cond)) +
  geom_density() +
  ggtitle("Density Plots (dat)")
```

# (e) Density plots with semitransparent fill
```{r}
ggplot(dat, aes(x = rating, fill = cond)) +
  geom_density(alpha = 0.3) +
  ggtitle("Density with Transparency (dat)")
```

# (f) Read "diabetes_train.csv" into 'diabetes' and 
#     apply same functions for 'mass' attribute, grouping by 'class'
```{r}
diabetes <- read.csv("diabetes_train.csv", header = TRUE)
str(diabetes)  # Check structure to see the column names
```

# Overlaid histograms for 'mass' by 'class'
```{r}
ggplot(diabetes, aes(x = mass, fill = class)) +
  geom_histogram(binwidth = 1, alpha = 0.5, position = "identity") +
  ggtitle("Overlaid Histograms (diabetes)")
```


# Interleaved histograms
```{r}
ggplot(diabetes, aes(x = mass, fill = class)) +
  geom_histogram(binwidth = 1, position = "dodge") +
  ggtitle("Interleaved Histograms (diabetes)")
```


# Density plots
```{r}
ggplot(diabetes, aes(x = mass, colour = class)) +
  geom_density() +
  ggtitle("Density Plots (diabetes)")
```


# Density plots with semitransparent fill
```{r}
ggplot(diabetes, aes(x = mass, fill = class)) +
  geom_density(alpha = 0.3) +
  ggtitle("Density with Transparency (diabetes)")
```



############################################################
# 4. Read "titanic.csv" and do base R operations 
############################################################

```{r}
passengers <- read.csv("titanic.csv", header = TRUE)
```


# (a) Drop NA rows and get summary
```{r}
passengers_noNA <- na.omit(passengers)  # Removes rows with any NA

summary(passengers_noNA)
```


# (b) Filter (Sex == "male") using base R subset
```{r}
passengers_male <- subset(passengers, Sex == "male")
# Print first rows (optional)
head(passengers_male)
```


# (c) Arrange in descending order by Fare
```{r}
passengers_descFare <- passengers[order(-passengers$Fare), ]
head(passengers_descFare)
```


# (d) Mutate: Add FamSize = Parch + SibSp
```{r}
passengers$FamSize <- passengers$Parch + passengers$SibSp
head(passengers)
```


# (e) Group by Sex, summarise mean(Fare) and sum(Survived)
#     We'll use tapply in base R.
```{r}
meanFare_bySex <- tapply(passengers$Fare, passengers$Sex, mean, na.rm = TRUE)
numSurv_bySex  <- tapply(passengers$Survived, passengers$Sex, sum, na.rm = TRUE)

# Combine results in a small data frame

groupedResults <- data.frame(
  Sex           = names(meanFare_bySex),
  meanFare      = as.numeric(meanFare_bySex),
  totalSurvived = as.numeric(numSurv_bySex)
)

groupedResults
```



############################################################
# 5. Quantiles of 'skin' in the diabetes data 
############################################################

We need the 10th, 30th, 50th, 60th percentiles
```{r}
skin_quantiles <- quantile(diabetes$skin, probs = c(0.1, 0.3, 0.5, 0.6), na.rm = TRUE)
cat("\nQuantiles (10%, 30%, 50%, 60%) of 'skin' in diabetes:\n")
print(skin_quantiles)
```